---
title: "Reproducible Research"
author: "Denford Madenyika"
date: "December 22, 2016"
output: html_document
---

## Variables included in the dataset

## Here we are going to global options and load libraries needed for this assignment

```{r}
library(lattice)
library(knitr)
opts_chunk$set(echo=TRUE,results="show",cache=TRUE)
setwd('C:/Users/denfordm/Desktop/Coursera/Reproducible_Research')
```
I will first load the data

Process/transform the data (if necessary) into a format suitable for your analysis

```{r}
dfMaster <- read.table("activity.csv",
               header=TRUE,
               sep=",",
               stringsAsFactors = FALSE,
               colClasses = c("numeric","Date","numeric")
               )
```

```{r}
df <- dfMaster
```
 Find the total number of steps taken per day

Generate a histogram to show number of steps taken per day

Find the mean and median of the total number of steps taken per day

```{r}
TotalStepsByDay <- aggregate(df$steps,list(date=df$date),sum,na.rm=TRUE)
xAxisBreakdown = seq(from=0,to=25000,by=2000) 
hist(TotalStepsByDay$x,
      breaks = xAxisBreakdown,
      main="Total Steps per Day",
      col="blue",
      xlab="Steps",
      ylab="Days",
      xaxt="n")
axis(side=1,at=xAxisBreakdown,labels=xAxisBreakdown)
```

 Calculate the mean

```{r}
stepMean <- mean(TotalStepsByDay$x,na.rm=T)
print(paste("The Mean number of steps per day is",round(stepMean,1)))
```
 Calculate the median

```{r}
stepMedian <- median(TotalStepsByDay$x,na.rm=T)
print(paste("The Median number of steps per day is",round(stepMedian,1)))
```
 Calculate the average daily activity pattern through the time series function

 I am interested in the hours

```{r}
intHours <- df$interval %/% 100
```

 I am only interested in the 24 hour format, two digit hours
```{r}
intHours <- ifelse(intHours < 10,paste("0",intHours,sep=""),intHours)
```
 Here I am interested in the minutes

```{r}
intMinutes <- df$interval %% 100
```

 I also just want two digit minutes
```{r}
intMinutes <- ifelse(intMinutes < 10,paste("0",intMinutes,sep=""),intMinutes)
```

 I concatenate minutes and hours and convert to a time with strptime()
```{r}
intTime <- paste(intHours,":",intMinutes,sep="")
intTime <- strptime(intTime,format="%H:%M")
```

 Then I add time variable into the data set

```{r}
df <- cbind(df,intTime)
```

 This gives me the time function
 I use it to compute number of steps

```{r}
SPI <- aggregate(df$steps,list(intTime=df$intTime),mean,na.rm=TRUE)
plot(SPI$intTime,SPI$x,
     type = "l",
     col = "blue",
     main = "Average Steps per Interval",
     xlab = "Interval",
     ylab = "Average Steps")
```
 I am also interested in the interval with highest interval

```{r}
MaxStepAvg <- max(SPI$x)
IntervalWithMaxStepAvg <- SPI$intTime[SPI$x == MaxStepAvg]
```

Here I will handle missing values


```{r}
countNAs <- sum(is.na(df$steps))
```

Average Interval Steps

```{r}
names(SPI)[names(SPI)=="x"] <- "avgIntervalSteps"
```

  average by intTime

```{r}
dfWithAvg <- merge(x=df,y=SPI,by="intTime",all.x=TRUE)
```


Use steps, if available, else avgIntervalSteps.
```{r}
dfWithAvg$imputedSteps <- ifelse(is.na(dfWithAvg$steps), 
                                 dfWithAvg$avgIntervalSteps,
                                 dfWithAvg$steps)
```

 Average

```{r}
TotalStepsByDayImputed <- aggregate(dfWithAvg$imputedSteps,list(date=dfWithAvg$date),sum,na.rm=TRUE)
xAxisBreakdown = seq(from=0,to=25000,by=2000) 
hist(TotalStepsByDayImputed$x,
      breaks = xAxisBreakdown,
      main="Frequency of Total Steps per Day",
      col="blue",
      xlab="Steps",
      ylab="Days",
      xaxt="n")
axis(side=1,at=xAxisBreakdown,labels=xAxisBreakdown)
```

 Mean of imputed steps

```{r}
stepMeanImputed <- mean(TotalStepsByDayImputed$x,na.rm=T)
print(paste("Mean number of imputed steps per day is",round(stepMeanImputed,1)))
```


Median number of imputed steps

```{r}
stepMedianImputed <- median(TotalStepsByDayImputed$x,na.rm=T)
print(paste("Median number of imputed steps per day is",round(stepMedianImputed,1)))
```

Weekdays averages Weekend averages I will compute and compare the averages

```{r}
dfWithAvg$weekday <- weekdays(dfWithAvg$date)
dfWithAvg$weekendFlag <- ifelse(dfWithAvg$weekday=="Saturday" | dfWithAvg$weekday=="Sunday","Weekend","Weekday")
```


Average steps per day for weekend and weekdays

```{r}
SPI2 <- aggregate(dfWithAvg$imputedSteps,list(intTime=dfWithAvg$intTime,weekendFlag=dfWithAvg$weekendFlag),mean,na.rm=TRUE)
```

Set up a sequence

```{r}
xn <- seq(min(dfWithAvg$intTime),max(dfWithAvg$intTime),by="4 hour")
```

24 hour format

```{r}
xyplot(x~intTime|weekendFlag,
  data=SPI2,
  type="l",
  layout=c(1,2),
  xlab = "Time Interval (24-hour clock)",
  ylab = "Average Steps",
  main = "Avg Steps per Day - Weekend vs  Weekday",
       scales=list(
                  x=list(
                        at=xn,
                        labels=format(xn,"%H:%M")
                        )
                  )
        )

```




